Combinatorics of the geometry of Wilson loop diagrams I: equivalence classes via matroids and polytopes

### 2019-07-10 GWLDCI.tex
read to beginning of 3.1

As described in the introduction, Wilson loop diagrams come from the study of SYM $N = 4$ in the physics literature \st{and therefore} \hlfix{where each propagator corresponds to a section of  a $k$-line bundle over twistor space. Thus}{I feel the strike out was too vague. The highlighted may be too technical. Can someone find baby bear's text?} each diagram $W$ is associated to a matrix of data $C(W)$, which we now describe. \sanote{Susama ended here}

@@

The matrix $C(W)$ will allow us to associate a matroid to each Wilson loop diagram. We first give a quick summary of the matroid terminology that we will need; it is not intended as a comprehensive introduction to matroids and the interested reader is referred to \cite{OxleyMatroidBook}

@@

Finally, we describe several important types of matroids. A matroid of rank $k$ with a ground set of size $n$ is called {\em realizable} if there exists some $A$ in the $k\times n$ real Grassmannian $\mathbb{G}_{\mathbb{R}}(k,n)$ whose non-zero $k\times k$ minors are exactly those with columns indexed by an element of $\cB$. If all the maximal minors of the matrix are nonnegative then the matroid realized by this matrix is called a {\em positroid}.  In other words a positroid is a matroid\footnote{Note that matroid isomorphism allows arbitrary permutations of the ground set, while positroid isomorphism only allows cyclic permutations in order to preserve the nonnegativity of the minors.  In particular, the property of being a positroid is not preserved under matroid isomorphism.} that is realized by an element of the totally nonnegative Grassmannian $\Gr(k,n)$. Finally, a {\em uniform matroid} of rank $r$ is a matroid in which any set of size $\leq r$ is independent.

@@

@@

Throughout, we take the {\em matroid defined by $W$} to be the matroid $M(W)$ of Theorem \ref{thm WLD defines matroid}. Note from Definition~\ref{def:CWmatrix} that the vertices of the diagram $W$ correspond to columns of the associated matrix $C(W)$. Therefore, for any subset of vertices $V \subseteq [n]$, we can consider the restriction $C(W)|_V$ of $C(W)$ consisting of the columns corresponding to $V$. Since the non-zero entries of $C(W)$ are independent real variables, \hlfix{the matrix $C(W)|_V$ does not have full rank if and only if there are fewer than $|V|$ rows with non-zero entries.}{I don't think this is quite true?  I agree, it could for instance have a column of 0s} This is equivalent to saying that $V$ is a set of dependent columns of $C(W)$ if and only if $|\Prop(V)| < |V|$. Comparing this to Theorem \ref{thm:WLDmatroid}, we see that $M(W)$ is also the matroid realized by $C(W)$.

@@

Let $W = (\cP,[n])$ be an admissible Wilson loop diagram, and $M(W)$ its associated matroid. Where it will not cause confusion we conflate the two objects, identifying vertices of $W$ with elements of the ground set $[n]$ in $M(W)$.

@@

To prove (4), we need to show that $F(P)$ is maximally dependent. If $F(P) = [n]$ then this is automatic, so suppose not and let $v \in [n] \setminus F(P)$. In other words, $v \in V(P^c)$ and so $v$ supports some propagator $q \not\in P$.  Let $S \subseteq F(P)$ be an independent set of maximal size. Then $\Prop(S) \subseteq P$ (from the definition of $F(P)$) and no subset of $S$ supports fewer propagators than the number of vertices it contains (by Theorem~\ref{thm WLD defines matroid} this is one characterization of an independent set in $M(W)$). Since $v$ supports a new propagator $q \not\in P$, the set $S \cup \{v\} \subseteq F(P) \cup\{v\}$ also satisfies this independence condition. Thus $\rk(F(P) \cup\{v\}) = \rk(F(P)) + 1$, as required.
\end{proof}

@@

>>>>

As described in the introduction, Wilson loop diagrams come from the study of SYM $N = 4$ in the physics literature \st{and therefore} \hlfix{where each propagator corresponds to a section of  a $k$ dimesional vector bundle over twistor space. Thus}{I feel the strike out was too vague. The highlighted may be too technical. Can someone find baby bear's text?} each diagram $W$ is associated to a matrix of data $C(W)$, which we now describe. \sanote{Susama ended here}

@@

The matrix $C(W)$ allows us to associate a matroid to each Wilson loop diagram. We first give a quick summary of the matroid terminology that we will need; it is not intended as a comprehensive introduction to matroids and the interested reader is referred to \cite{OxleyMatroidBook}

@@

Finally, we describe several important types of matroids. A matroid of rank $k$ with a ground set of size $n$ is called {\em realizable} if there exists some $A$ in the $k\times n$ real Grassmannian $\mathbb{G}_{\mathbb{R}}(k,n)$ whose non-zero $k\times k$ minors are exactly those with columns indexed by an element of $\cB$. If all the maximal minors of the matrix are nonnegative then the matroid realized by this matrix is called a {\em positroid}.  \sanote{we seem to have eliminated all other reference to positroid in this paper. Shoudl we get rid of this? I would prefer to keep this, and say that admissible Wilson loop diagrams define positroids, but leave it there. For me, the point of this paper is to elucidate the map from WLDs to matroids, showing what we already know about the image is part of that.} In other words a positroid is a matroid\footnote{Note that matroid isomorphism allows arbitrary permutations of the ground set, while positroid isomorphism only allows cyclic permutations in order to preserve the nonnegativity of the minors.  In particular, the property of being a positroid is not preserved under matroid isomorphism.} that is realized by an element of the totally nonnegative Grassmannian $\Gr(k,n)$. Finally, a {\em uniform matroid} of rank $r$ is a matroid in which any set of size $\leq r$ is independent.

@@

\sanote{Everything from here applies to weakly admissible. In fact, I think the dfn of admissible in my paper with Amat is what we call weakly admissible here. Should we change to weakly?}

@@

Throughout, we take the {\em matroid defined by $W$} to be the matroid $M(W)$ of Theorem \ref{thm WLD defines matroid}. Note from Definition~\ref{def:CWmatrix} that the vertices of the diagram $W$ correspond to columns of the associated matrix $C(W)$. Therefore, for any subset of vertices $V \subseteq [n]$, we can consider the restriction $C(W)|_V$ of $C(W)$ consisting of the columns corresponding to $V$. Since the non-zero entries of $C(W)$ are independent real variables, any subset $V$ that has fewer than $|V|$ rows of non-zero entries does not have full rank. This is equivalent to saying that if $V$ is a set of columns of $C(W)$ with $|\Prop(V)| < |V|$, $V$ is a dependent set. Any set subset of columns containing such a $V$ cannot be independent. Comparing this to Theorem \ref{thm:WLDmatroid}, we see that $M(W)$ is also the matroid realized by $C(W)$. \sanote{fixed dependent set argument}

@@

Let $W = (\cP,[n])$ be an admissible Wilson loop diagram, and $M(W)$ its associated matroid. Where it will not cause confusion we conflate the two objects, we identify vertices of $W$ with elements of the ground set $[n]$ in $M(W)$.

@@

To prove (4), we need to show that $F(P)$ is maximally dependent. If $F(P) = [n]$ then this is automatic, so suppose not and let $v \in [n] \setminus F(P)$. In other words, $v \in V(P^c)$ and so $v$ supports some propagator $q \not\in P$.  Let $S \subseteq F(P)$ be an independent set of maximal size. Then $\Prop(S) \subseteq P$ (from the definition of $F(P)$) and no subset of $S$ supports fewer propagators than the number of vertices it contains (by Theorem~\ref{thm WLD defines matroid}) \st{this is one characterization of an independent set in $M(W)$)}. Since $v$ supports a new propagator $q \not\in P$, the set $S \cup \{v\} \subseteq F(P) \cup\{v\}$ also satisfies this independence condition. Furthermore, since $S \cup v$ is independent, by the first point in this lemma, $|S| = \rk(S) = \rk F(P)$. Similarly $|S \cup v| = |S|+1 = \rk (S \cup v) \leq \rk(F(P) \cup v)$.  Thus $\rk(F(P) \cup\{v\}) = \rk(F(P)) + 1$, as required. \sanote{tweaked}
\end{proof}

@@

This section explores the correspondence between Wilson loop diagrams and positroids, established in \cite{wilsonloop}. In that paper, the authors show that admissible Wilson loop diagrams define matroids that are also positroids, Theorem \ref{put it back in then ref}. Furthermore, they show that if two Wilson loop diagrams are equivalent, they define the same matroid [ref thm].

In this section, we take this analysis further, showing the converse of Theroem [ref] from loc. cit. (see Theorem \ref{same matroid iff equiv}. We also enumerate the number of diagrams that map to an particular matroid, Corollary \ref{number of equiv diagrams}. To do this, we establish a connectection between Wilson loop diagrams and chord diagrams, subsection \ref{sec: polygon partitions}, and prove some matroidal properties of exact subdiagrams in subsection \ref{sec: exact diagram matroidal props}  including that they correspond to uniform matroids (Theorem \ref{exactuniformthm}), and the associated matroid can be written as as a contraction by the complementary propagator flat (Theorem \ref{exact diagrams contractions}).\sanote{stopped here}\sanote{need to discuss positroids, and whether positroids should appear earlier so that they can be discussed here.}
